---
title: "Basic Concept of Statistics"
author: "Paolo Girardi and Livio Finos"
date: "8/10/2020"
output:
  pdf_document:
    highlight: haddock
    latex_engine: xelatex
    toc: yes
  html_document:
    df_print: paged
    toc: yes
---
\newpage
\begin{center}
Ph.D. Course in Neuroscience   
Calendar of the Basic Courses – Academic Year 2020-2021   
Basic Concept of Statistics   
\  
Lesson 2 - Optional and preliminary course on use of R  
\end{center}

# Descriptive Statistics

## Univariate Statistical Analysis with R  

We import the dataset test.csv  

```{r}
test<-read.csv("test.csv",sep=";",header=T,dec=",")
head(test) #the first 6 rows
str(test)
summary(test)
```

### Analysis of the Age variable  

```{r}
# A histogram with the density plot
hist(test$Age,prob=T)
lines(density(test$Age),col=2)
```
The distribution is skewed. A qqplot can be used to visualise the distribution 
```{r}
qqnorm(test$Age)
qqline(test$Age,col=2,lty=2)
```
The graph reports the comparison between the theoretical quantile of a Normal distribution and quantiles of the variabile Age. If the points follow the red line, a normal distribution can be assumed.  

The function \textit{boxplot()} 
```{r}
boxplot(test$Age)
```

### Analysis of the BMI variable  

```{r}
# A histogram with the density plot
hist(test$BMI,prob=T)
lines(density(test$BMI),col=2)
```
The distribution looks simmetric but there is the presence of outliers (values to low and to high respect to the central cloud)
```{r}
qqnorm(test$BMI)
qqline(test$BMI,col=2,lty=2)
```
The QQplot confirms the presence of anomalous values of BMI.     
A unique plot with many boxplots.  
```{r}
par(mfrow=c(1,3)) # 1  row 3 cols
boxplot(test$Age,xlab="Age")
boxplot(test$BMI,xlab="BMI")
boxplot(test$ACT,xlab="ACT")
par(mfrow=c(1,1))
```

### Analysis of the Education variable  

```{r}
# A barplot with the density plot
plot(test$Education)
```

The barplot reports the frequency of each modality of the categorical variable.
But in this variable there is an order.
So we define the order as follows  

```{r}
#here the levels
levels(test$Education)
test$Education<-factor(test$Education,levels=
c("lower primary","primary","upper primary",
"secondary","degree","post-degree"),ordered =TRUE)
plot(test$Education) # here is ordered
```

Here a pie plot

```{r}
pie(table(test$Education))
```

The function \textit{table()} permits to obtain a frequency table

```{r}
table(test$Education)
#or a relative frequency table with the function prop.table()
prop.table(table(test$Education))
```

## Bivariate Statistical Analysis with R 
The dataset reported the results of 150 subjects on ACT e SAT tests.  
Some variables influences the performances.

### Quantitative vs qualitative variables
```{r}
#ACT vs Gender and Education
boxplot(test$ACT~test$Gender)
boxplot(test$ACT~test$Education)
```

```{r}
#SATV vs Gender and Education
boxplot(test$SATV~test$Gender)
boxplot(test$SATV~test$Education)
```


```{r}
#SATQ vs Gender and Education
boxplot(test$SATQ~test$Gender)
boxplot(test$SATQ~test$Education)
```


### Quantitative vs Quantitative variables

```{r}
#ACT vs Age
plot(test$Age,test$ACT)
cor(test$Age,test$ACT) # pearson
cor(test$Age,test$ACT,method="spearman") # spearman
```

```{r}
#ACT vs BMI
plot(test$BMI,test$ACT)
cor(test$BMI,test$ACT,method="spearman") # spearman
```
```{r}
#ACT vs Stress
plot(test$Stress,test$ACT)
cor(test$Stress,test$ACT,method="spearman") # spearman
```
