---
title: "Basic Concept of Statistics"
author: "Paolo Girardi and Livio Finos"
date: "15/10/2020"
output:
  pdf_document:
    highlight: haddock
    latex_engine: xelatex
    toc: yes
---
\newpage

Ph.D. Course in Neuroscience    
Calendar of the Basic Courses – Academic Year 2020-2021    
Basic Concept of Statistics    
\  
Lesson 2 - Optional and preliminary course on use of R   


# Descriptive Statistics
In R some useful functions for the descriptive analysis are:  
\begin{tabular}{ll}
- plot(x, y): & bivariate plot of x (on the x-axis) and y (on the y-axis);\\
- hist(x): & histogram of the frequencies of x\\
- barplot(x): & histogram of the values of x; use horiz=FALSE for horizontal bars\\
- dotchart(x): & if x is a data frame, plots a Cleveland dot plot (stacked plots line-by- line and column-by-column)\\
- pie(x): & circular pie-chart\\
- boxplot(x): & box-and-whiskers plot\\
- stripplot(x): & plot of the values of x on a line (an alternative to boxplot() for small sample sizes)\\
- mosaicplot(x):& mosaic plot from frequencies in a contingency table\\
- qqnorm(x): & quantiles of x with respect to the values expected under a normal law\\
\end{tabular}

## Univariate Statistical Analysis with R  

We import the dataset test.csv   

```{r}
# we set the Work Directory
setwd("/Users/Paolo/Dropbox/Dottorato_Neurosciences/2020_2021")
# import the test.csv file
test<-read.csv("test.csv",sep=";",header=T,dec=",")
head(test) #the first 6 rows
str(test)
summary(test)
```

This dataset is formed by the first 150 subjects of a larger dataset.
The dataset reported some information about the SAT and ACT test, performed on people during some job's selection.

### Analysis of the Age variable  
 
```{r}
# A histogram with the density plot
summary(test$Age)
hist(test$Age,prob=T)
lines(density(test$Age),col=2)
```

The distribution is skewed, in particular few numbers after 40 years old. A qqplot can be used to visualise the distribution 
```{r}
qqnorm(test$Age)
qqline(test$Age,col=2,lty=2)
```
The graph reports the comparison between the theoretical quantile of a Normal distribution and quantiles of the variabile Age. If the points follow the red line, a normal distribution can be assumed.   


The function \textit{boxplot()} performs (box and whiskers plot) as follows  


```{r}
boxplot(test$Age)
```

### Analysis of the BMI variable  
The BMI (Body Mass Index) is the ratio between weight/height.  


```{r}
# A histogram with the density plot
summary(test$BMI)
hist(test$BMI,prob=T)
lines(density(test$BMI),col=2)
```

The distribution looks simmetric but there is the presence of outliers (values to low and to high respect to the central cloud).  

```{r}
qqnorm(test$BMI)
qqline(test$BMI,col=2,lty=2)
```
The QQplot confirms the presence of anomalous values of BMI.     


A unique plot with many boxplots.  

```{r}
par(mfrow=c(1,3)) # 1  row 3 cols
boxplot(test$Age,xlab="Age")
boxplot(test$BMI,xlab="BMI")
boxplot(test$ACT,xlab="ACT")
par(mfrow=c(1,1))
```

### Analysis of the Education variable  

```{r}
# A barplot with the frequency
barplot(table(test$Education))
```

The barplot reports the frequency of each modality of the categorical variable.
But this variable has an order. So we define the order as follows:  

```{r}
#here the levels
levels(test$Education)
test$Education<-factor(test$Education,levels=
c("lower primary","primary","upper primary",
"secondary","degree","post-degree"),ordered =TRUE)
plot(test$Education) # here is ordered
```

Here a pie plot  

```{r}
pie(table(test$Education))
```

The function \textit{table()} permits to obtain a frequency table  

```{r}
table(test$Education)
#or a relative frequency table with the function prop.table()
prop.table(table(test$Education))
```
### Analysis of the Stress variable  

The variable Stress is an integer values expressed on a likert scale (the common question:"How much are you stressed from 1 to 6?"). The likert scale is not numeric (variable on ratio scale), but it is an ordinal variable.  
With the command \textit{factor()} R can set a factor, a categorical variable, even if it is formed by numbers.

```{r}
is(test$Stress)
test$Stress<-factor(test$Stress)
table(test$Stress)
#or a relative frequency table with the function prop.table()
prop.table(table(test$Stress))
# the same for the variable social 
test$Social<-factor(test$Social)
```

## Bivariate Statistical Analysis with R 
The dataset reported the results of 150 subjects on ACT e SAT tests.  
Some variables influences the performances.  

We try to reply to the question: "What are the factors that influenced the ACT, SATV and SATQ test?"  


### Quantitative vs qualitative variables
```{r}
#ACT vs Gender and Education
boxplot(test$ACT~test$Gender)
#change colour with col argument and labels
boxplot(test$ACT~test$Gender,col=c("red","green"),ylab="ACT",xlab="Gender")
boxplot(test$ACT~test$Education,ylab="ACT",xlab="Education")
boxplot(test$ACT~test$Stress,ylab="ACT",xlab="Stress")
boxplot(test$ACT~test$Social,ylab="ACT",xlab="Social")

#all the plot in a unique figure
par(mfrow=c(2,2))
boxplot(test$ACT~test$Gender,col=c("red","green"),ylab="ACT",xlab="Gender")
boxplot(test$ACT~test$Education,ylab="ACT",xlab="Education")
boxplot(test$ACT~test$Stress,ylab="ACT",xlab="Stress")
boxplot(test$ACT~test$Social,ylab="ACT",xlab="Social")
par(mfrow=c(1,1))
```

Here the statistical analysis for SATV and SATQ.    
What are the comments on these charts?  

```{r}
#SATV vs Gender and Education, Stress and Social
boxplot(test$SATV~test$Gender,col=c("red","green"),ylab="SATV",xlab="Gender")
boxplot(test$SATV~test$Education,ylab="SATV",xlab="Education")
boxplot(test$SATV~test$Stress,ylab="SATV",xlab="Stress")
boxplot(test$SATV~test$Social,ylab="SATV",xlab="Social")
```


```{r}
#SATQ vs Gender and Education, Stress and Social
boxplot(test$SATQ~test$Gender,col=c("red","green"),ylab="SATQ",xlab="Gender")
boxplot(test$SATQ~test$Education,ylab="SATQ",xlab="Education")
boxplot(test$SATQ~test$Stress,ylab="SATQ",xlab="Stress")
boxplot(test$SATQ~test$Social,ylab="SATQ",xlab="Social")
```

### Quantitative vs Quantitative variables

Analysis of ACT vs. Age and BMI
```{r}
#ACT vs Age
plot(test$Age,test$ACT)
```

I can use some graphical parameter to have a better graph (please see ?plot or ?par for more options) :  

- main, sub: title and subtitle 
- xlab, ylab: label of the x and y axis  
- xlim, ylim: limits of the x and y axis  
- type: type of plot  
- lty: type of lines  
- pch:plot symbol  
- cex: scale factor  
- col: color of points etc.  
```{r}
#ACT vs Age
plot(test$Age,test$ACT,pch=16,col="green",xlab="Age",ylab="ACT")
```

Pearson correlation   

```{r}
cor(test$Age,test$ACT) # pearson
cor(test$Age,test$ACT,method="spearman") # spearman
plot(test$BMI,test$ACT)
cor(test$BMI,test$ACT,method="spearman") # spearman
```

```{r}
par(mfrow=c(1,2))
plot(test$Age,test$SATV)
plot(test$BMI,test$SATV)
par(mfrow=c(1,1))
```

```{r}
par(mfrow=c(1,2))
plot(test$Age,test$SATQ)
plot(test$BMI,test$SATQ)
par(mfrow=c(1,1))
```


Is there a correlation between test scores (ACT, SATQ, SATV)? 


```{r}
plot(test[,c("ACT","SATV","SATQ")])
cor(test[,c("ACT","SATV","SATQ")])
cor(test[,c("ACT","SATV","SATQ")],method="spearman")
```
