---
title: "Basic Concept of Statistics"
author: "Paolo Girardi and Livio Finos"
date: "22/10/2020"
output:
  pdf_document:
  highlight: haddock
latex_engine: xelatex
toc: yes
---
\newpage

Ph.D. Course in Neuroscience    
Calendar of the Basic Courses – Academic Year 2020-2021   
Basic Concept of Statistics

Lesson 4 - Introduction to testing 






# Hypothesis testing  

## Before we start (in R)  

```{r}
#clean the memory
rm (list=ls ())
ECHO=FALSE
# customize the output of knitr
knitr :: opts_chunk$set (fig.align="center")#, fig.width=6, fig.height=6)

```

## Diving reflex test - Dataset 

The diving reflex, also known as the diving response and mammalian diving reflex, is a set of physiological responses to immersion. One of the main outcomes is a heart rate reduction. The effect is enhanced by the water temperature.

We measure the reduction in the heart rate of 10 subjects (children) for different temperature of the water (in Fahrenheit degree).

The values of `Temperature` are in Fahrenheit degree. 
The values of `Reduction` are in beat per second.

(data are fictitious)

To read the data

```{r}
setwd("/Users/Paolo/Dropbox/Dottorato_Neurosciences/2020_2021")
diving<-read.csv("diving_reflex.csv")
str(diving)
```

---

We plot the data

```{r,fig.height=4,fig.width=4}
plot(x=diving$Temperature,y=diving$Reduction,pch=20,col=2,cex=2)
```

# Measures of Dependence and the Simple linear model
## Measuring the dependence

we define:

- $X=Temperature$  
- $Y=Reduction$

We review some famous index to measure the (linear) dependence among two variables

## Covariance and Variance

**Covariance** between $X$ and $Y$:

$\sigma_{xy}=\frac{\sum_{i=1} ^ n (x_i- \bar{x}) (y_i- \bar{y} )}{n}$

- values between $- \infty$ and $\infty$  
- $\sigma_{xy} \approx 0$: there is no dependency between $X$ and $Y$  
- $\sigma_{xy} >> (<<) 0$: there is a strong positive (negative) dependency between $X$ and $Y$

## Correlation
With the Covariance it is difficult to understand when the relationship between $X$ and $Y$ is strong / weak.
We note that

$- \sigma_{x} \sigma_{y} \leq \sigma_{xy} \leq \sigma_{x} \sigma_{y}$
<!-- we divide each memeber by $\sigma_x \sigma_y$: -->
is quivalent to
$-1 \leq \frac{\sigma_{xy}}{\sigma_{x} \sigma_{y}} \leq 1$


**Correlation** between $X$ and $Y$:

$\rho_{xy}=\frac{\sigma{xy}}{\sigma_{x} \sigma_{y}} =
\frac{\sum_{i=1} ^ n (x_i- \bar{x}) (y_i- \bar{y})}{\sqrt{\sum_{i=1} ^ n (x_i- \bar{ x}) ^ 2} \sqrt{\sum_{i=1} ^ n (y_i- \bar{y}) ^ 2}}$

- values between $-1$ and $1$
- $\rho_{xy} \approx 0$: there is no dependency between $X$ and $Y$
- $\rho_{xy} \approx 1 (-1)$: there is a strong positive (negative) dependency between $X$ and $Y$


## Linear Trend, the least squares method
We describe the relationship between   
`Reduction` and
`Temperature` with a straight line.

$Reduction \approx \beta_0 + \beta_1 Temperature$  
$Y=\beta_0 + \beta_1X$

Let's draw a line 'in the middle' of the data.

--- 

The **least-squares estimator**

We look for the one that passes more 'in the middle', the one that minimizes the sum of the squares of the residues:

$\hat{\beta}_0$ and $\hat{\beta}_1$ such that  
$\sum_{i=1} ^ n (y_i - (\hat{\beta}_0 + \hat{\beta}_1x_i )) ^ 2$
is minimum.

---

```{r,echo=FALSE}
model=lm(Reduction~Temperature,data=diving)
```

Estimates:  

- Angular coefficient: $\hat{\beta}_1=\frac{\sigma_{xy}}{\sigma_{xx}}=\rho_{xy}\frac{\sigma_{y}}{\sigma_{x}}=\frac{\sum_{i=1}^n(x_i- \bar{x})(y_i-\bar{y})}{\sum_{i=1}^n (x_i-\bar{x})^2}=$ `r coefficients(model)[2]`  
- Intercept: $\hat{\beta}_0=\bar{y}-\hat{\beta}_1\bar{x}=$ `r coefficients(model)[1]`
- Response (estimated $y$): $\hat{y}_i=\hat{\beta}_0 + \hat{\beta}_1x_i$
- Residuals (from the estimated response):
$y_i - (\hat{\beta}_0 + \hat{\beta}_1 x_i)=y_i- \hat{y}_i$


and therefore the least squares are the sum of the squared residuals:
$\sum_{i=1} ^ n (y_i- \hat{\beta}_0 - \hat{\beta}_1x_i) ^ 2=\sum_{i=1} ^ n (y_i- \hat{y}_i ) ^ 2$

---

A graphical representation:

```{r}
# lm() is the function to estimate a classical linear model in R
model=lm(Reduction~Temperature,data=diving)
coefficients(model)
```

---

```{r}
plot(diving$Temperature,diving$Reduction,pch=20,col=2,cex=1)
coeff=round(coefficients(model),1)
title(paste("Y=",coeff[1],"+",coeff[2],"*X"))
abline(model,col=1)
``` 

## Interpretation of the coefficients

- $\beta_0$ indicates the value of $y$ when $x=0$ (where the line intersects the ordinate axis).
- $\beta_1$ indicates how much $y$ grows or decreases as a unit of $x$ grows
    - If $\beta_1=0$ there is no relation between $x$ and $y$.$Y$is constant (horizontal ), knowing$x$does not change the estimate of $y$
    - If $\beta_1> (<) 0$ the relation between $x$ and $y$ is positive (negative). When $X$ passes from $x$ a $x + 1$ the estimate of $Y$ changes from $\hat{y}$ to $\hat{y} + \hat{\beta}_1$


# Permutation approach to Hypothesis Testing

## Some remarks

Let's note that all the measures above does not make any assumptions on the random process that generate them.

Let's assume that $Y$ - and possibly $X$ - is not fix, while it is generated by a random variable.

--- 

The question:
**Is there a relationship between $Y$ and $X$**?

We estimated 
$\hat{\beta}_1=$ `r coefficients(model)[2]`

but the **true value** $\beta_1$ is really different from 0 (i.e. no relationship)?  
Otherwise, is the distance to 0 is due to the random sampling?

- **Null Hypothesis** $H_0: \ \beta_1=0$ (the **true** $\beta_1$, not its estimate $\hat{\beta}_1$!).
There is no relationship between $X$ and $Y$.

- **Alternative Hypothesis **$H_1: \ \beta_1 <0$
The relationship is negative

Other possible specifications of $H_1: \ \beta_1> 0$ and, more commonly,  $H_1: \ \beta_1 \neq 0$.


## Permutation tests - in a nutshell
As a toy example, let use a sub-set of the data of 3 subjects:
```{r, echo=FALSE,fig.height=4,fig.width=4}
set.seed(1)
(reactionREd=diving[sample(nrow(diving),3),])
plot (reactionREd$Temperature, reactionREd$Reduction, pch=20, col=1,xlab="Temperature",ylab="Reduction")
abline (lm (Reduction ~ Temperature, data=reactionREd))
```

---

- *If $H_0$* is true: there is no linear relationship between $X$ and $Y$
- Therefore, the trend observed on the data is due to chance.
- Any other match of $x_i$ and $y_i$ was equally likely to occur
- I can generate the datasets of other hypothetical experiments by exchanging the order of the observations in $Y$.
- How many equally likely datasets could I get with$X$ and $Y$observed?
$3 * 2 * 1=3!=6$ possible datasets.

Remark: Here we only assume that $y$ is a random variable. The only assumption here is the exchangeability of the observations: the joint density $f(y_1,\ldots,y_n)$ does not change when the ordering of $y_1,\ldots,y_n$ is changed.

## All potential datasets
```{r, echo=FALSE}
make_perm_plot<-function(yperm=sample(diving$Temperature),divingPerm=diving){
  divingPerm$Temperature=yperm
  model=lm(Reduction~Temperature,data=divingPerm)
  plot(divingPerm$Temperature,divingPerm$Reduction,pch=20,col=3,cex=2,xlab="Temperature",ylab="Reduction")
  coeff=round(coefficients(model),2)
  title(paste("Y=",coeff[1],"+",coeff[2],"*X"))
  abline(model,col=2)
}
Y=cbind (reactionREd[,3],reactionREd[c(1,3,2),3],reactionREd[c(2,1,3),3],reactionREd[3:1,3],reactionREd[c(2,3,1),3],reactionREd[c(3,1,2),3])
X=reactionREd$Temperature
# Y=cbind (c (2,1,4), c (1,2,4), c (1,4,2),
# c (4,1,2), c (4,2,1), c (2,4,1)) * 10
# X=1: 3
# cbind (X=X, Y)
par (mfrow=c (2,3))
for (i in 1: 6){
make_perm_plot(Y[,i],divingPerm = reactionREd)
}
par (mfrow=c (1,1))
```

---

## Random permutations

In our data set, if we apply the same principle...

How many permutations of the vector $y_1,\ldots,y_n$ are possible? $n!=10!=3628800$.

big, perhaps not too big ... but what happen with, for example, $n=20$? We got $20!=2.432902e+18$. This is too big, definitely! 

We calculate a smaller (but sufficiently large) $B$ of random permutations. 

here some example

---

**`Temperature` vs a permutations of `Reduction`**

```{r, results='asis', echo=FALSE}
par(mfrow=c(2,3))
for(i in 1:6)
  make_perm_plot()
par(mfrow=c(1,1))
```

---

We repeat `r (B=10000)` times and we look at the histogram of the $\hat{\beta}_1$

```{r}
# beta_1 estimated on the observed data:
beta1=coefficients(lm(Reduction~Temperature,data=diving))[2]
# function that permutes the y values and calculates the coeff beta_1
my.beta.perm <- function(Y,X){
  model=lm(sample(Y)~X)
  coefficients(model)[2]
}
#replicate it B-1 times
beta.perm= replicate(B,my.beta.perm(diving$Reduction, diving$Temperature ))
```

----

```{r,echo=FALSE}
#the observed dataset is one of the possible permutations
beta.perm=c(beta1,beta.perm)
# str(beta.perm)
hist(beta.perm,50)
points(beta1,0,lwd=3,col=1)
```

## How likely $\hat{\beta}_1 ^{obs}$ was?
(before the experiment!)

How likely was it to get a $\leq \hat{\beta}_1 ^{obs}$ value among the many possible values of $\hat{\beta}_1 ^{*b}$ (obtained by permuting data)?


Remarks:

- $\hat{\beta}_1 ^{* b}< \hat{\beta}_1 ^{obs}$ (closer to 0): less evidence against $H_1$ than $\hat{\beta}_1 ^{obs}$
- $\hat{\beta}_1 ^{* b} \geq \hat{\beta}_1 ^{obs}$: equal or more evidence towards $H_1$ than $\hat{\beta}_1 ^{obs}$


## Calculation of the p-value

Over B=`r B` permutations we got `r sum (beta.perm <= beta1)` times a 
$\hat{\beta}_1 ^{* b} \leq \hat{\beta}_1 ^{obs}$.

The p-value (significance) is
$p=\frac{\# (\hat{\beta}_1 ^{* b} \geq \hat{\beta}_1 ^{obs})}{B + 1} =$ `r (p=sum(beta.perm<= beta1)/B)`
In our example is
```{r}

mean((beta.perm<=beta1))

```

## Interpretation

The probability of $p=P (\hat{\beta}_1 ^ * \leq \hat{\beta}_1=$ `r round (coefficients (model), 3)[2]` $| H_0)$ is equal to $p =$ `r p`, i.e. very small.  
So, it was unlikely to get a value like this **IF $H_0$ is true**.

Neyman-Pearson's approach has made common the use of a significance threshold for example $\alpha=.05$ (or $=. 01$).
When $p \leq \alpha$ rejects the hypothesis that there is no relationship between X and Y ($H_0$). If so, we are inclined to think that $H_1$ is true (there is a positive relationship).

- Type I error: False Positive  
the true hypo is $H_0$ (null correlation), BUT we accept $H_1$ (correlation is positive)
- Type II error: False Negative  
the true hypo is $H_1$ (positive correlation), BUT we do not reject $H_0$ (null correlation)

**Type I error control**
  
We want to guarantee not to get false relationships (a few false positives), better to be conservative. To make this, we want to bound the probability to make a false discovery:
  
$P (p-value \leq \alpha | H_0) \leq \alpha$
  
We built a machinery that in the long run (many replicates of the experiment) finds false correlations with probability $\alpha$ (e.g. $0.05=5\%$).


## Composite alternatives (bilateral)
The hypothesis $H_1: \beta_1 <0$ (the relation is negative) must be justified with a priori knowledge.

More frequently, the Alternative hypothesis is appropriate:
$H_1: \beta_1 \neq 0$
(there is a relationship, I do not assume the direction)

I consider anomalous coefficients estimated as very small but also very large ('far from 0').
The p-value is 
$p=\frac{\#(|\hat{\beta}_1^{*b} | \geq|\hat{\beta}_1^{obs}|)}{B+1}=$ `r (mean(beta.perm<beta1)+mean(beta.perm>-beta1))`

In our example
```{r}
mean(beta.perm<beta1)+mean(beta.perm>-beta1)
```


## Some remarks

- Do not be confused with bootstrap methods. The former are extractions without reintegration, the latter with. The former have almost optimal properties and have (almost always) an exact control of the first type errors.
- A general approach and are applicable in many contexts. Very few assumptions.
- Some dedicated R packages: 
- [flip](http://cran.r-project.org/web/packages/flip/index.html) (the development version is on [github](https://github.com/livioivil/flip))
- [coin](http://cran.r-project.org/web/packages/coin/index.html) 
- [permuco](https://cran.r-project.org/web/packages/permuco/index.html)   

- They are of limited applicability when there are many variables involved.

# Parametric Linear Model
## From permutation tests (nonparametric) to parametric tests
We can see that the histogram of the statistical tests (calculated on the permuted data) is well described by a **Gaussian **(normal) curve.

----

```{r}
hist(beta.perm,50,probability=TRUE,col=2)
curve(dnorm(x,mean(beta.perm),sd(beta.perm)),add=TRUE,col=1,lwd=3)
points(beta1,0,lwd=3,col=1)
```

## The (simple) linear model
We assume that the observed values are distributed around true values
$\beta_0 + \beta_1 X$ according to a Gaussian law:

$Y=\textrm{linear part} + \textrm{normal error}$

$Y=\beta_0 + \beta_1 X + \varepsilon$

**Assumptions of the linear model **

- the $\boldsymbol{y_i=\beta_0 + \beta_1 x_i + \varepsilon_i}$
the relationship between $X$ and the true (mean) $Y$ is linear.
-  the **observations** are **independent** each others (knowing the value of the $y_i$ observation does not help me to predict the value of $y_{i + 1}$). The random part is $\varepsilon_i$, these are the independent terms.
- $\boldsymbol{\varepsilon_i \sim N (0, \sigma ^ 2), \ \forall i=1, \ldots, n}$ errors have normal distribution with zero mean and common variance (homoschedasticity: same variance).



## Hypothesis testing
If these assumptions are true,

$\hat{\beta_1} \sim N (\beta_1, \sigma ^ 2 / \sum (x_i- \bar{x}) ^ 2)$

We calculate the test statistic:

$t=\frac{\hat{\beta_1}}{std.dev\ \hat{\beta_1}}=\frac{\hat{\beta_1}}{\sqrt{\sum_{i=1} ^ n (y_i- \hat{y}_i) ^ 2 / \sum (x_i- \bar{x}) ^ 2 / (n-2)}}$

If $H_0: \beta_1=0$, $t \sim t (n-2)$ is true

On `reaction` data and $H_1: \beta_1 \neq 0$ (bilateral alternative)


----

```{r}
model=lm (Reduction ~ Temperature, data=diving)
summary(model) $ coefficients
```

Similar result, but much more assumptions!


## Power is nothing without control
We don't know if the data are generated  under $H_0$ or under $H_1$. 

But we have a tool (the test) that  

- if the data are generated **under $H_0$**: it suggests (wrong) $H_1$ (i.e. $p\leq \alpha$, type I error, false positive) with probability $\alpha$. E.g. $\alpha=.05$, low probability.
- if the data are generated **under $H_1$**: it suggests (correct) $H_1$ (i.e. true positive) with probability larger than $\alpha$.  
This is the Power of a test.
The Power is unknown, but we hope it is as high as possible.

## Terminology

- Probability of **Type I error** (Probability of **False Positive**, $\alpha$): the probability to find a relationship when it does not exist (true $H_0$, the test judges $H_1$).
- Probability of **Type II error** (Probability of **False Negative**): the probability NOT to find a relationship when it does exist (true $H_1$, the test judges $H_0$).


- **Specificity**: the probability NOT to find a relationship when it does NOT exist (true $H_0$, the test judges $H_0$). it is equal to 1 - Type I error.
- **Power** (**Sensitivity**): the probability to find a relationship when it does exist (true $H_1$, the test judges $H_1$). it is equal to 1 - Type II error.

## Properties

If the parametric assumptions are valid, the test guarantes  

- the control of the type I error at the $\alpha$ level,  
- the maximum power (minimum error of type II $\beta$) among all the possible tests,  
- asymptotic consistency (if they are under $H_1$ rejection always for sufficiently large $n$).

The permutation tests usually have slightly less power and converge to the corresponding parametric tests, IF they exist.

# Descriptive Analysis and Test for Longitudinal Data

## Local Field Potentials DATASET
The data are recordings of Local Field Potentials from an array with 220 channels inserted in the rat barrel cortex. 
The intracerebral local field potential (LFP) is a measure of brain activity that reflects the highly dynamic flow of information across neural networks.

The dataset contains 200 rows (# of channels) and 1000 columns (first 1000 time steps).

```{r}
setwd("/Users/Paolo/Dropbox/Dottorato_Neurosciences/dataset")
lcp<-read.csv("lfpreduced_ex.csv",header=TRUE)
lcp<-as.matrix(lcp)
```

## Univariate Analisys 
There are 2 dimensions, channels and time 
### by channels  

```{r}
channels_mean<-apply(lcp,1,mean,na.m=T)
channels_var<-apply(lcp,1,mean,na.m=T)
summary(channels_mean)
hist(channels_mean)
boxplot(channels_mean)
```

### by time  

```{r}
time_mean<-apply(lcp,2,mean,na.m=T)
time_var<-apply(lcp,2,var,na.rm=T)
summary(time_mean)
plot(time_mean)
plot(time_mean,type="l")
abline(h=0,lty=2)
```

### plot all the timeseries  

```{r}
plot(time_mean,ylim=range(lcp),type="n")
for(i in 1:200){
lines(lcp[i,],lwd=0.3,col="grey",lty=2)
}
lines(time_mean)
```


### correlation between sites  

```{r}
library(corrplot) # install.packages("corrplot")
# we take the first 30 sites
corrplot(cor(t(lcp[1:30,])))
## time series 16 contains missing data
```

### cross-correlation at different lag  

```{r}
ccf(lcp[1,],lcp[2,])
ccf(lcp[1,],lcp[3,])
```

## A simple test
There is some difference between the first 500 time points and the second 500 time points?  
A measure of brain activity is the range of the measured LFP.  
We calculate the range (max-min) for each channel.  

```{r}
t1<-as.vector(diff(apply(lcp[,1:500],1,range)))
t2<-as.vector(diff(apply(lcp[,501:1000],1,range)))
```

The distribution of the measured range in t1 (1-500 timepoints) and in t2  (501-1000 timepoints) is reported below

```{r}
par(mfrow=c(1,2))
boxplot(t1,main="T1",ylab="Range")
boxplot(t2,main="T2",ylab="Range")
par(mfrow=c(1,1))
boxplot(t1-t2,main="T1-T2",ylab="Range")
shapiro.test(t1)
shapiro.test(t2)
```

Shapiro-Wilk normality test reported a non normal distribution for the range of the values of the channels.
We assume a spatial independence between sites (is not true... but it is necessary to perform the test); a non parametric test (Wilcoxon test) is performed below  

```{r}
wilcox.test(t1,t2,paired = TRUE)
```

There is a strong difference ($p \approx 0$) in the range of the first 500 timepoints respect to the second 500 timepoints.